---
title: "Assingment_D"
author: "George Acostalemus"
format: html
editor: visual
---

```{r, warning = FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)
library(zoo)
library(tibble)
library(readr)
```

```{r, warning=FALSE, message=FALSE}
# Define the station ID
station_id <- "bhbm3"

# Create file root and tail
file_root <- paste0("https://www.ndbc.noaa.gov/view_text_file.php?filename=", station_id, "h")
tail <- ".txt.gz&dir=data/historical/stdmet/"

# Function to load a single year's buoy data
load_buoy_data1 <- function(year) {
  path <- paste0(file_root, year, tail)
  
  message("Loading year: ", year)
  
  if (year < 2007) {
    # For older files, header is on first line and read.table works better
    header <- scan(path, what = 'character', nlines = 1, quiet = TRUE)
    
    buoy <- read.table(path, fill = TRUE, header = TRUE, sep = "")
    
    # Some older files are missing mm (minutes) and TIDE columns — add them
    if (!"mm" %in% names(buoy)) {
      buoy <- add_column(buoy, mm = NA, .after = "hh")
    }
    if (!"TIDE" %in% names(buoy)) {
      buoy <- add_column(buoy, TIDE = NA, .after = "VIS")
    }
    
  } else {
    # For newer files, fread is faster and more reliable
    header <- scan(path, what = 'character', nlines = 1, quiet = TRUE)
    
    buoy <- fread(path, header = FALSE, skip = 1, fill = TRUE)
    
    # Assign column names from the header
    setnames(buoy, header)
  }
  
  # Return the data frame
  return(buoy)
}

```

```{r}
load_buoy_data <- function(year) {
  path <- paste0(file_root, year, tail)

  header <- scan(path, what = 'character', nlines = 1)
  num_columns <- length(header)  

  if (num_columns == 16) {
    buoy <- read.table(path, fill = TRUE, header = TRUE, sep = "")
    buoy <- add_column(buoy, mm = NA, .after = "hh")
    buoy <- add_column(buoy, TIDE = NA, .after = "VIS")

  } else if (num_columns == 17) {
    buoy <- read.table(path, fill = TRUE, header = TRUE, sep = "")
    buoy <- add_column(buoy, TIDE = NA, .after = "VIS")

  } else {
    buoy <- fread(path, header = FALSE, skip = 1, fill = TRUE)
    setnames(buoy, header)
  }

  return(buoy)
}
all_data <- lapply(2005:2024, load_buoy_data)
combined_data <- rbindlist(all_data, fill = TRUE)
```

```{r}
# Load and Combine Multiple Years Was getting errors so chatpGPT helped me fixed column namings

# Step 1: Convert year columns to character if they exist
combined_data <- combined_data %>%
  mutate(
    YY = if ("YY" %in% names(.) ) as.character(YY) else NA_character_,
    `#YY` = if ("#YY" %in% names(.) ) as.character(`#YY`) else NA_character_,
    YYYY = if ("YYYY" %in% names(.) ) as.character(YYYY) else NA_character_
  )

# Step 2: Create a single year column using coalesce (prioritize full year YYYY)
combined_data <- combined_data %>%
  mutate(Year = coalesce(YYYY, `#YY`, YY))

# Step 3: Convert other date/time columns to character and then numeric safely
combined_data <- combined_data %>%
  mutate(
    MM = as.character(MM),
    DD = as.character(DD),
    hh = as.character(hh),
    mm = if ("mm" %in% names(.)) as.character(mm) else "00"
  )

# Step 4: Create datetime using lubridate ymd_hm or ymd_h depending on availability of minutes
combined_data <- combined_data %>%
  mutate(
    datetime_str = paste(Year, MM, DD, hh, mm, sep = "-"),
    datetime = ymd_hm(datetime_str, quiet = TRUE) # use ymd_h if no minutes data
  )

# Step 5: Remove intermediate columns if desired (optional)
combined_data <- combined_data %>% select(-YY, -`#YY`, -YYYY, -datetime_str)

# Step 6: Replace flagged missing values (e.g., 99, 999, 9999) with NA across numeric columns
combined_data <- combined_data %>%
  mutate(across(where(is.character), as.numeric)) %>%
  mutate(across(where(is.numeric), ~na_if(., 99))) %>%
  mutate(across(where(is.numeric), ~na_if(., 999))) %>%
  mutate(across(where(is.numeric), ~na_if(., 9999)))

```

# Question: How does air temperature change throughout the year? Any trends from 2020–2024? ?

This question focuses on air temperature use real measurements, encourages monthly or seasonal grouping, and spans multiple years.

To visualize this, we will use line plots to see if there's any trends within the air temperature.

```{r}
# Combine multiple years into one dataset
temp_data <- combined_data %>%
  filter(!is.na(datetime),
         Year >= "2020", Year <= "2024",
         !is.na(ATMP)) %>%  # assuming air temp is in column "ATMP"
  mutate(
    Year = as.integer(Year),
    Month = month(datetime, label = TRUE)  # Month as factor with labels Jan, Feb...
  )

# Calculate average monthly temperature by year
monthly_avg_temp <- temp_data %>%
  group_by(Year, Month) %>%
  summarise(Avg_Temp = mean(ATMP, na.rm = TRUE)) %>%
  ungroup()

# Plot
ggplot(monthly_avg_temp, aes(x = Month, y = Avg_Temp, color = factor(Year), group = Year)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Average Monthly Air Temperature (2020–2024)",
    x = "Month",
    y = "Average Air Temperature (°C)",
    color = "Year"
  ) +
  theme_minimal()
```

Each line represents one year from 2020-2024 The temperature follows a clear season cycle whee the lowest temperature are in January and the highest temperature are in July. Each line is shaped like a bell curve which is consistent year over year. These findings could point to broader long-term environmental changes, such as global warming.
