---
title: "HOMEWORK D"
author: "Georg Acostalemus"
format: 
   html:
      embed-resources: true
editor: visual
---

## Assignments due on Monday

1.  please submit github URL for the buoy reading assignment

2.  Exercises from R4DS\
    a - 3.2.5: 1,4,5\
    b - 3.3.5: 1,4\
    c - 3.5.7: 1,2,4,6

## Cheat Sheets

[dplyr](https://dplyr.tidyverse.org/)

[tidyr](https://tidyr.tidyverse.org/)

[stringr](https://stringr.tidyverse.org/)

### using a cheatsheet

Examples from dplyr

## R4DS Chapter 3

```{r}
#| label: R4DS_ch_3-1
#| warning: true
#| message: true

library(nycflights13)
library(tidyverse)

flights_here <- flights

```

### Rows - Manipulate Cases

filter() arrange() distinct()

```{r}
#| label: rows
#| warning: false
#| message: false

flights |> filter(dep_delay > 120)

delay_120 <- flights_here |> filter(dep_delay > 120)

library(magrittr) 

flights_here %<>% filter(dep_delay > 120)

delay_order <- flights_here |> arrange(dep_delay)


```

### Columns -- Manipulate Variables

select() mutate()

```{r}
#| label: columns

 flights |> select(year, month, day)
 
 m1 <- flights |> 
     mutate(
      gain = dep_delay - arr_delay,
      speed = distance / air_time * 60
      )

```

### Summarize and groups

```{r}
#| label: groups

flights |> 
   group_by(month)

flights |> 
   group_by(month) |>
   summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE)
   )

```

```{r}
# Problem 1
# Arrival delays of 2 hours or more
flights |> 
  filter(arr_delay >= 2)

# Flights destination to Houston 
flights |> 
  filter(dest == "IAH" | dest == "HOU")

# Flights operated by Delta, United or American
flights |> 
  filter(carrier == "AA" | carrier == "DL" | carrier == "UA")

# Flights departed in summer
flights |>
  filter(month %in% c(7,8,9))

# More than two hours late but didn't leave late
flights |>
  filter(arr_delay > 120, dep_delay <= 0)

# Delayed by at least an hour, but made up over 30 minutes in flight
flights |>
  filter(dep_delay >= 60, (dep_delay - arr_delay) >30)
```

3.2.5 #4

```{r}
# Was there a flight everyday in 2013

flights_per_day <- flights %>% 
  group_by(year, month, day) %>% 
  summarise(flight_count = n())

days_with_flights <- nrow(flights_per_day)
days_with_flights == 365 # Check if there's a flight every day

# We got true, there was a flight everyday in 2013
```

3.2.5 \# 5

```{r}
# FLights with the longest and shortest distance

# Longest Distance
flights |>
  filter(distance == max(distance, na.rm = TRUE))
# 4983 miles is the longest distance flight

# Shortest Distance
flights |>
  filter(distance == min(distance, na.rm = TRUE))
# 17 miles is the shortest distance flight

```

3.3.5 \# 1

```{r}
flights_check <- flights %>%
  filter(!is.na(dep_time) & !is.na(sched_dep_time)) %>%
  mutate(
    sched_dep_mins = (sched_dep_time %/% 100) * 60 + (sched_dep_time %% 100),
    dep_mins = (dep_time %/% 100) * 60 + (dep_time %% 100),
    calculated_delay = dep_mins - sched_dep_mins,
    delay_diff = dep_delay - calculated_delay
  ) %>%
  filter(delay_diff != 0)

# View rows where the calculated delay doesn't match dep_delay
flights_check

# Code was made with chatBot 
# Explanation below
```

Here we subtracted the actual time departure minus the scheduled time departure. If the difference in delay is 0 in all or most cases then the relationship is holding well.

3.3.5 \# 4

The any function is used when selecting columns from a data frame. With the vector variables \<- c("year", "month", "day", "dep_delay", "arr_delay") because if one of the columns doesn't exist in our data then we would usually get an error. With any_of function it'll select any columns that are present and skips the missing ones, exceuting the code line by line. Here is the code to further prove this.

```{r}
vector <- c("year", "month", "day", "nonexisttent_column")

flights |>
  select(any_of(vector))
# You will see three out the four columns. year, month and day
```

3.5.7 \# 1

```{r}
# Calculate the average arrival day per carrier
carrier_delays <- flights %>%
  group_by(carrier) %>%
  summarise(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    n_flights = n()
  ) %>%
  arrange(desc(avg_arr_delay))

carrier_delays

# The highest avg_arr_delay is the worst. 
# In this case Frontier (F9) airlines has the worst average delays
```

3.5.7 #2

```{r}
# Flights that are most delayed upon departure from each destination
max_dep_delays <- flights %>% 
  group_by(dest) %>% 
  summarise(max_dep_delays = max(dep_delay, na.rm = TRUE))

most_delayed_flights <- flights %>%
  inner_join(max_dep_delays, by = "dest") %>%
  filter(dep_delay == max_dep_delays) %>%
  arrange(desc(dep_delay))  # optional: see the worst at top
# Chat bot



most_delayed_flights %>%
  select(dest, year, month, day, carrier, flight, origin, dep_delay)
# It seems that JFK airport in NYC has long departure delays.
```

3.5.7 #4

Using a negative n in slice_min() will drop the n largest or n smallest values rather than running the code.

```{r}
# Example with a different datset
df <- tibble(x = c(1, 3, 5, 7, 9))

# Negative n: drop 2 smallest values
slice_min(df, x, n = -2)
# The highest two numbers get dropped

```

3.5.7 \# 6

```{r}
df <- tibble(
  x = 1:5,
  y = c("a", "b", "a", "a", "b"),
  z = c("K", "K", "L", "L", "K")
)

# a
df |> # group_by groups y columns treat rows with the same y value
  group_by(y)

# b
df |> 
  arrange(y) # Sort the rows by column y

# c
df |>
  group_by(y) |>
  summarize(mean_x = mean(x)) # Gets the mean of x with every y group
# First it groups by y then takes the average of x in each grou

# d
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x)) # We get one row for each y and z pair
# This code keeps the first group of y unless you telll it otherwise.

# e
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x), .groups = "drop") # almost similar to part d

# .groups = "drop"tells dplyr to remove al grouping after summarise. unlike part d
# where it was still grouped by y

df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x)) # It will summarize the average of x

df |>
  group_by(y, z) |>
  mutate(mean_x = mean(x)) # it will mutate the average of x into a new column

# The difference is summarize collaspses each group to one row
# Mutate keeps original rows and adds summary value in a new column.
```
